\begin{acknowledgments}

I would like to express my heartfelt gratitude to:

\begin{itemize}

\item
my teachers and peers from IIT Bombay, for being a never-ending 
source of inspiration and for giving my career the best start 
that I could have hoped for.

\item
Somesh Jha, for six years of patient advice and support, for
ensuring me a productive research environment, and for shaping 
me as a researcher.

\item
Trent Jaeger, for being a wonderful mentor and for being
unconditionally supportive of all my efforts.

\item
Trishul Chilimbi, Thomas Reps, Sanjit Seshia, Michael Swift, 
and all my collaborators over the years for giving me the 
opportunity to work with and learn from them.

\item
Trent Jaeger, Susan Horwitz, Marvin Solomon, Michael Swift and
Parameswaran Ramanathan, for serving on my thesis committee.

\item
my academic siblings and other members of the security research group 
at UW-Madison, for making the group a productive and enjoyable one to 
do research in.

\item
all my friends over the years, both at UW-Madison and at IIT Bombay, 
for support, encouragement, and for keeping me sane.

\end{itemize}

Thank you very much! 

My biggest source of advice, encouragement, motivation and support 
are undoubtedly Amma and Appa, and mere words will do injustice to 
express my gratitude to them. Their guidance and sacrifice made this
possible, and their affection and belief in me made it all worthwhile.
My success is also theirs. To them, I dedicate this work.

\paragraph{Funding.} My graduate studies were funded by ONR grants 
N00014-01-1-0708 and N00014-01-1-0796 and NSF grant CNS-0448476. 
The Secure Systems Department at IBM Thomas J. Watson Research Center 
(Hawthorne, NY) and the Runtime Analysis and Design Group at Microsoft 
Research (Redmond, WA) employed me as a summer intern in Summer 2005 
and Summer 2004, respectively. Their support is gratefully acknowledged.

\end{acknowledgments}
